Demo for a Quality of Service networking with Software Defined Network.

### 2017-03-15 report/paper.tex
other corrections on english. end to read after, now class

\subsection{Adaptability of the network}
An adaptable Network has to real-time reroute traffic in order to ensure a lot of demands.
A general efficient Network has to manage some tasks, for example avoiding links congestion, reroute traffic when some malfunction accure on a link
and running a link path algorithm. 
In addition our network has to be adaptable in order to ensure a QoS type of service.

To reach this gools we needed to have a global knowledge on the network and we had to modify the routing table on every routerboard with a
global purpose.
The performance of a single connection and entire network can ensure various type of metrics.
In order to ensure the traffic shaping based on network informations, there is a need of an automatically updatable structure,
where the link statistics can be saved to allow the best path for a data stream. 
The best metric will be described below.

\subsection{Exterior comunication}
Our network needed to be able to communicate with the external internet to be useful for a demo which demonstrate the capabilities of a SDN network.
The first thing we needed to accomplish was to make the controller aware of what switch and which port of the switch is connected to the external internet:
to do so, we allowed our controller to be able to receive and process in-line parameters. We provided two parametes to the controller, so that
it could recognize where the traffic to the internet need to be sent. Theese two parameters were the mac address of the
port of a switch that is connected to the link and the network address of our network, to let the controller know if a request from a host was meant to another host
within the network or the data stream was intended to the internet. We provided only one gateway, as the Routeboard had few ports available for the packet switching
and only four SDN-capable switches to be able to simulate a real network that was generating traffic.

@@

\section{Results}\label{sec:results}
\subsection{Controller}
We developed a controller that can handle a mesh network with loops, avoiding
packets flooding. The controller stores the switch and the connection between them and knows which hosts are connected to each switch port starting from the hypothesys that only one
host is connected to each port, unless the port is the one connected to the internet.
With the use of statistics, the controller gets the status of the network and, if necessary,
tries to adjust the traffic and to put in the link with the higher path loss the torrent traffic.
On the other hand it sets up the connection for two gaming hosts going through a link with less delay and tries to redistribute all the 
traffic in a way that all the nodes have an average throughput as lowest as possible.
\\
\newline
We have used some principal functions released with POX controller:
\begin{itemize}
 \item \texttt{pox.openflow.discovery.launch()} this function runs an algorithm in order to discover the network. Every time a new
 topological change occurs this function returns an event on POX with which we can update our topological knowledge.
 It does not provide the host event\cite{pox}.
 \item \texttt{pox.openflow.spanning_tree.launch()} this function is necessary with a presence of loop on topology
 because with a spanning_tree algorithm it avoids the broadcast storm of messages.

@@

  The links on our network logically have many parameters in witch we can storage network's performance; in practice a NetworkX's graph support only 
  one or two link's attributes so we have create a lots of graphs with the same topological information but different link's attributescd .

@@

\subsection{Traffic's Analysis}\label{subsec:traf}
We used laptop connected via
GigabitEthernet interface and Ethernet cable cat.5a to receive the traffic from the destination host.
Our laptop has been connected via Wi-Fi to our access point via protocol 802.11n.
No other traffic sources were active during the testing, so the router was handling only
the traffic from laptop when sniffing traffic. The connection used was an ADSL2+ via copper.
We did 3 type of traffic analysis: Gaming traffic, Skype® and torrent traffic. We analyzed
\#Mb for the gaming traffic, \#Mb for the VoIP traffic and 3GB for the torrent traffic.
The analysis involved the average number of packets per second, the packet size, the protocol (TCP or UDP).

@@

\subsubsection{Delay}
Monitoring round-trip time provides important insights for network troubleshooting and traffic engineering.
The  common  monitoring  technique  is  to  actively  send  probe packets  from  selected  vantage  points  hosts  or  middleboxes.
In traditional  networks, the control over the network routing is
limited,  making  it  impossible  to  monitor  every  selected  path. The  emerging  concept  of  Software  Defined  Networking  simplifies
network  control.  However,  OpenFlow,  the  common  SDN protocol,   does   not   support   RTT   monitoring   as   part   of   its
specification. In this experience we have tryed to emply an tecnic to escape this problem\cite{rtt}.
\newline The idea is to measure the time that occure from sending a packet from controller to a switch and the time in which the packet return from
an other switch.
\\
\newline First we have to discovery round trip time from controller to every single switch on our topology.
In order to do this on the controller we start a timer, we send a proper packet to the switch specifing the \texttt{OFPP_CONTROLLER - Send to the controller}
option and when the message return to the controller we stop the timer and calculare RTT for the switch.
\begin{figure}[!h]
 \centering
 \includegraphics[scale=0.70]{images/rtt2.png}
 \caption{\emph{RTT between controller and OF switch}}
 \label{fig:topo}
\end{figure}
\newline
If we need to calculate the delay on a specific link, the rtt has to calculate with the headed switch of this link.
\\
\newline
Now we create a proper message with unique id. number, let's start a timer \texttt{TM1} and then send this message to a first switch specifing the output port.
Then when the packet comes to the second switch it identify the particular message and sends to the controller that stops it's dedicated timer \texttt{TM2}.
\newline This part required a particolar flow entry on the second switch in order to send to the controller only some particulare packet.
\begin{figure}[!h]
 \centering
 \includegraphics[scale=0.70]{images/rtt1.png}
 \caption{\emph{Pck from controller to itself following a particular way.}}
 \label{fig:topo}
\end{figure}
\\
\newline At this point we can calculate the delay of the link.
\\
\\
 delay = (TM2 - TM1) - RTT1/2 - RTT2/2
\\
\\
Teorethically this method must works but in practice there are a lot of delay introduced from computational congestion on the virtualized controller.
What we observe is a estimated delay like a few seconds which is not the case, infact if we made some \texttt{ping test} from one host to another host
in this SDN network we realize delay around 3ms.
\\
\\
The following picture shows the visual output that we can obtain from delay's graph. In this graph we storage only the core information;
\begin{figure}[!h]
 \centering
 \includegraphics[scale=0.26]{images/delay.png}
 \caption{\emph{Delay's graph.}}
 \label{fig:topo}
\end{figure}

@@

	The first data that was collected was the actual link capacity, which depends both on the cable used to connect
	different devices and the effective capacity for each ethernet port linked by the cable itself. Everytime a device supporting
	openflow connects to the controller, it gives to the latter information about its status and its capabilities. One of them is
	the link capacity in three ways: the first is the \textit{supported} capacity, which is the phisical maximum capacity, the second is the
	\textit{advised} capacity, which name describes itself and the latter is the \textit{current} one. We choose to save the last one,
	as it gives the most useful about the link state. This information, moreover, is the one that non-SDN switches use to choose the \textit{metric}
	parameter. The simulation with Mininet showed us that this type of information can be used, but the Routeboards didn't support it.
	As a matter of fact, we weren't able to exploit this capacity in our network realization.
		\\
		\\
  \subsubsection{Throughput}

	Using the flow statistics, we analyzed the traffic going through a link and, with a simple division, we got the normalized throughput.
	We exploited the information given by the switches with the match structure: this means that we used the throughput for the TCP traffic,
	the one for the UDP and for each of them we could know toward which IP the traffic is directed or is from. This was exploited with mininet
	simulations, but not in the set up in the laboratory. With the latter, we used the statistics as an indicator of the bitrate of a link
	and divided the host to categories based on it.
\\
\subsubsection{Packet error rate}
The port statistics given by the switches contains different informations about the state of each port in a switch. To classify a link for its
reliability, we divided the number of packets with erroneous symbols that couldn't be corrected with a FEC by the total number of packets that were sent,
getting the packet error rate. The choice we made when exploiting this statistic was privileging the gaming and normal traffic, wherease the
heavy traffic was directed to links with highest packet error rate.  As the switches exchanges a packet before recognizing most of the error, both because the
phisical error are corrected using blocks coding and other error are found using a \textit{Frame Check Sequence} like CRC, the byte error were considered worse as
those are indirect measures themself and the error derived from the calculations could grow up.

@@

For each statistic we got and each parameter we obtained, we changed the weight of the graphs. Every graph we used had a specific weight, so the
Dijkstra algorithm for a weighted graph gave different minimum path between two hosts. From the parameter we saved, we choose to change the
route to reach the destination and viceversa. The traffic follow the same path in both the directions and depends only on the type of traffic:
we choose to set the minimum unweighted path if no statistics were available, the minimum throughput path for the standard traffic, the minimum
delay for the gaming one and the maximum packet error rate for the heavy traffic generated by a host.
Two threshold were chosen based on the traffic analysis described on Sec. \ref{subsec:traf} which are 200 bytes/s for the gaming traffic and one megabyte/s
when the node is considered in heavy traffic. Those two threshold were made as they are average values and considering that our network is made of few really fast link,
so on one side the lower limit is sufficient not to consider gaming traffic to be a normal traffic, wherease it's easy to reach the high level of traffic load between two machines.
The changing of the path were made only if the last change happened before one minute because the torrent traffic has, in its protocol, a discovery
algortithm. This means that if a lot of changes were made, most of the traffic generated by the peer to peer, is the control one and the result is that the
load in the network will be heavy and last longer.

\subsection{ARP solution}
During our analysis and in our tests, we prove that in a SDN mesh network, the ARP packets are useless.
This type of packets is useful for a multipoint to multipoint connection where the connection is set between a
HUB and a set of computers or between a bridge and a computer. The first case is no more in use in common networks as
it higher the collision range. The bridges limits their collision range to the port. With openflow each switching unit can
act as a bridge or as a router. In both cases analyses the header of the L2 packet L3 packet and some of the L4 header.
If it has the rule to forward the packet, it will act as declared in the most specific matching rule. Otherwise it will
send to the controller a PacketIn and the controller will act as programmed. If an host want to communicate to another host whithin
the network, it sends the packet to the line and then the switch will check in its tables and execute the actions of the match found.
For an IP network, it will check the path to reach a destination and then send to the defined output port. The mac address in this case
became useless, as the connection is point to point. So, knowing the destination layer2 address is useless as the switch can act
regardless of it. The only thing the switch need to know is the layer2, layer 3 addresses and, at most, the layer 4 address.
This can be seen by a reader as a violation of the layers in the ISO/OSI stack, but to exploit the potentials of the SDN approach,
this is necessary.

@@

\subsection{Ip switching in pox}
During our tests, both with Mininet and in the lab, we found out that the core of POX, when a PacketIn event occoures, switches the source and the destination addresses,
both MAC and IP, if one of the hosts is already known. We weren't able to understand in which point of the code of POX  and why it happens.
% One of the problem encountered was that, if an host is already found by pox and there is a PacketIn event, the core of the controller
% switches the source and the destination IP and MAC address with the source one.
The problem was partially solved, checking out the
parameters known by the controller, i.e. if the two host are connected to two different switches and the controller switches the source with the
destination address, we can switch it back; the result is that the PacketIn event now has correct informations.
If the two host are in the same switch, the problem is solved using a controlled packet flooding, described in section \ref{sec:connectivity}.

\section{Conclusions}\label{sec:conclusion}
% Conclusions are a superbrief summary of what has been done and highlighting of the "take home message"
% ehi
For this course project, the report shows how the demo was made, in particular what problem we encountered
and what we exploited, showing what protocols could be unused in the future. We showed a possible simple data structure
that can allow the controller to be able to know the network, its topology and which route to set, based on the traffic.
Morover, we showed how to exploit the potentialities of the SDN network, giving particular importance on the
statistics and the path switching capabilities.
The multi-graph solution allowed us to connect the network and, based on its condition, choose which weight to apply
in order to give a Quality of Service based network. With some boards that supports the newer version of Openflow
(i.e. 1.5, the latest), the code can be ported, so it can be used with the Ryu core controller and more statistic can
be exploited. Moreover, the controller can set more strictive rules for the flow control and can get more informations from the
switches.

>>>>

\subsection{Adaptability of the network}
An adaptable Network has to real-time reroute traffic in order to ensure a lot of demands.
A general efficient Network has to manage some tasks, for example avoiding links congestion, reroute traffic when some malfunction occure on a link
and running a link path algorithm. 
In addition our network has to be adaptable in order to ensure a QoS type of service.

To reach this goals we needed to have a global knowledge on the network and we had to modify the routing table on every routerboard with a
global purpose.
The performance of a single connection and entire network can ensure various type of metrics.
In order to ensure the traffic shaping based on network informations, there is a need of an automatically updatable structure,
where the link statistics can be saved to allow the best path for a data stream. 
The best metric will be described below.

\subsection{Exterior comunication}
Our network needed to be able to communicate with the external internet to be useful for a demo which demonstrate the capabilities of a SDN network.
The first thing we needed to accomplish was to make the controller aware of what switch and which port of the switch is connected to the external internet:
to do so, we allowed our controller to be able to receive and process in-line parameters. We provided two parameters to the controller, so that
it could recognize where the traffic to the internet need to be sent. These two parameters were the mac address of the
port of a switch that is connected to the link and the network address of our network, to let the controller know if a request from a host was meant to another host
within the network or the data stream was intended to the internet. We provided only one gateway, as the Routeboard had few ports available for the packet switching
and only four SDN-capable switches to be able to simulate a real network that was generating traffic.

@@

\section{Results}\label{sec:results}
\subsection{Controller}
We developed a controller that can handle a mesh network with loops, avoiding
packets flooding. The controller stores the switch and the connection between them and knows which hosts are connected to each switch port starting from the hypotheses that only one
host is connected to each port, unless the port is the one connected to the internet.
With the use of statistics, the controller gets the status of the network and, if necessary,
tries to adjust the traffic and to put in the link with the higher path loss the torrent traffic.
On the other hand it sets up the connection for two gaming hosts going through a link with less delay and tries to redistribute all the 
traffic in a way that all the nodes have an average throughput as lowest as possible.
\\
\newline
We have used some principal functions released with POX controller:
\begin{itemize}
 \item \texttt{pox.openflow.discovery.launch()} this function runs an algorithm in order to discover the network. Every time a new
 topological change occurs this function returns an event on POX with which we can update our topological knowledge.
 It does not provide the host event\cite{pox}.
 \item \texttt{pox.openflow.spanning_tree.launch()} this function is necessary with a presence of loop on topology
 because with a spanning_tree algorithm it avoids the broadcast storm of messages.

@@

  The links on our network logically have many parameters in witch we can storage network's performance; in practice a NetworkX's graph support only 
  one or two link's attributes so we have create a lots of graphs with the same topological information but different link's attributes .

@@

\subsection{Traffic's Analysis}\label{subsec:traf}
We used laptop connected via
GigabitEthernet interface and Ethernet cable cat.5a to receive the traffic from the destination host.
Our laptop has been connected via Wi-Fi to our access point via protocol 802.11n.
No other traffic sources were active during the testing, so the router was handling only
the traffic from laptop when sniffing traffic. The connection used was an ADSL2+ via copper.
We did 3 type of traffic analysis: Gaming traffic, Skype® and torrent traffic. We analyzed
\#Mb for the gaming traffic, 70 Mb for the VoIP traffic and 3GB for the torrent traffic.
The analysis involved the average number of packets per second, the packet size, the protocol (TCP or UDP).

@@

\subsubsection{Delay}
Monitoring round-trip time provides important insights for network troubleshooting and traffic engineering.
The  common  monitoring  technique  is  to  actively  send  probe packets  from  selected  vantage  points  hosts  or  middleboxes.
In traditional  networks, the control over the network routing is
limited,  making  it  impossible  to  monitor  every  selected  path. The  emerging  concept  of  Software  Defined  Networking  simplifies
network  control.  However,  OpenFlow,  the  common  SDN protocol,   does   not   support   RTT   monitoring   as   part   of   its
specification. In this experience we have tryed to emply an tecnic to escape this problem\cite{rtt}.
\newline The idea is to measure the time that occur from sending a packet from controller to a switch and the time in which the packet return from
an other switch.
\\
\newline First we have to discovery round trip time from controller to every single switch on our topology.
In order to do this on the controller we start a timer, we send a proper packet to the switch specifying the \texttt{OFPP_CONTROLLER - Send to the controller}
option and when the message return to the controller we stop the timer and calculate RTT for the switch.
\begin{figure}[!h]
 \centering
 \includegraphics[scale=0.70]{images/rtt2.png}
 \caption{\emph{RTT between controller and OF switch}}
 \label{fig:topo}
\end{figure}
\newline
If we need to calculate the delay on a specific link, the rtt has to calculate with the headed switch of this link.
\\
\newline
Now we create a proper message with unique id. number, let's start a timer \texttt{TM1} and then send this message to a first switch specifying the output port.
Then when the packet comes to the second switch it identify the particular message and sends to the controller that stops it's dedicated timer \texttt{TM2}.
\newline This part required a particular flow entry on the second switch in order to send to the controller only some particular packet.
\begin{figure}[!h]
 \centering
 \includegraphics[scale=0.70]{images/rtt1.png}
 \caption{\emph{Pck from controller to itself following a particular way.}}
 \label{fig:topo}
\end{figure}
\\
\newline At this point we can calculate the delay of the link.
\\
\\
 delay = (TM2 - TM1) - RTT1/2 - RTT2/2
\\
\\
Theoretically this method must works but in practice there are a lot of delay introduced from computational congestion on the virtualized controller.
What we observe is a estimated delay like a few seconds which is not the case, in fact if we made some \texttt{ping test} from one host to another host
in this SDN network we realize delay around 3ms.
\\
\\
The following picture shows the visual output that we can obtain from delay's graph. In this graph we storage only the core information;
\begin{figure}[!h]
 \centering
 \includegraphics[scale=0.26]{images/delay.png}
 \caption{\emph{Delay's graph.}}
 \label{fig:topo}
\end{figure}

@@

	The first data that was collected was the actual link capacity, which depends both on the cable used to connect
	different devices and the effective capacity for each Ethernet port linked by the cable itself. Every time a device supporting
	openflow connects to the controller, it gives to the latter information about its status and its capabilities. One of them is
	the link capacity in three ways: the first is the \textit{supported} capacity, which is the physical maximum capacity, the second is the
	\textit{advised} capacity, which name describes itself and the latter is the \textit{current} one. We choose to save the last one,
	as it gives the most useful about the link state. This information, moreover, is the one that non-SDN switches use to choose the \textit{metric}
	parameter. The simulation with Mininet showed us that this type of information can be used, but the Routeboards didn't support it.
	As a matter of fact, we weren't able to exploit this capacity in our network realization.
		\\
		\\
  \subsubsection{Throughput}

	Using the flow statistics, we analyzed the traffic going through a link and, with a simple division, we got the normalized throughput.
	We exploited the information given by the switches with the match structure: this means that we used the throughput for the TCP traffic,
	the one for the UDP and for each of them we could know toward which IP the traffic is directed or is from. This was exploited with mininet
	simulations, but not in the set up in the laboratory. With the latter, we used the statistics as an indicator of the bitrate of a link
	and divided the host to categories based on it.
\\
\subsubsection{Packet error rate}
The port statistics given by the switches contains different information about the state of each port in a switch. To classify a link for its
reliability, we divided the number of packets with erroneous symbols that couldn't be corrected with a FEC by the total number of packets that were sent,
getting the packet error rate. The choice we made when exploiting this statistic was privileging the gaming and normal traffic, whereas the
heavy traffic was directed to links with highest packet error rate.  As the switches exchanges a packet before recognizing most of the error, both because the
physical error are corrected using blocks coding and other error are found using a \textit{Frame Check Sequence} like CRC, the byte error were considered worse as
those are indirect measures themselves and the error derived from the calculations could grow up.

@@

For each statistic we got and each parameter we obtained, we changed the weight of the graphs. Every graph we used had a specific weight, so the
Dijkstra algorithm for a weighted graph gave different minimum path between two hosts. From the parameter we saved, we choose to change the
route to reach the destination and viceversa. The traffic follow the same path in both the directions and depends only on the type of traffic:
we choose to set the minimum unweighted path if no statistics were available, the minimum throughput path for the standard traffic, the minimum
delay for the gaming one and the maximum packet error rate for the heavy traffic generated by a host.
Two threshold were chosen based on the traffic analysis described on Sec. \ref{subsec:traf} which are 200 bytes/s for the gaming traffic and one megabyte/s
when the node is considered in heavy traffic. Those two threshold were made as they are average values and considering that our network is made of few really fast link,
so on one side the lower limit is sufficient not to consider gaming traffic to be a normal traffic, whereas it's easy to reach the high level of traffic load between two machines.
The changing of the path were made only if the last change happened before one minute because the torrent traffic has, in its protocol, a discovery
algorithm. This means that if a lot of changes were made, most of the traffic generated by the peer to peer, is the control one and the result is that the
load in the network will be heavy and last longer.

\subsection{ARP solution}
During our analysis and in our tests, we prove that in a SDN mesh network, the ARP packets are useless.
This type of packets is useful for a multipoint to multipoint connection where the connection is set between a
HUB and a set of computers or between a bridge and a computer. The first case is no more in use in common networks as
it higher the collision range. The bridges limits their collision range to the port. With openflow each switching unit can
act as a bridge or as a router. In both cases analyses the header of the L2 packet L3 packet and some of the L4 header.
If it has the rule to forward the packet, it will act as declared in the most specific matching rule. Otherwise it will
send to the controller a PacketIn and the controller will act as programmed. If an host want to communicate to another host within
the network, it sends the packet to the line and then the switch will check in its tables and execute the actions of the match found.
For an IP network, it will check the path to reach a destination and then send to the defined output port. The mac address in this case
became useless, as the connection is point to point. So, knowing the destination layer2 address is useless as the switch can act
regardless of it. The only thing the switch need to know is the layer2, layer 3 addresses and, at most, the layer 4 address.
This can be seen by a reader as a violation of the layers in the ISO/OSI stack, but to exploit the potentials of the SDN approach,
this is necessary.

@@

\subsection{Ip switching in pox}
During our tests, both with Mininet and in the lab, we found out that the core of POX, when a PacketIn event occures, switches the source and the destination addresses,
both MAC and IP, if one of the hosts is already known. We weren't able to understand in which point of the code of POX  and why it happens.
% One of the problem encountered was that, if an host is already found by pox and there is a PacketIn event, the core of the controller
% switches the source and the destination IP and MAC address with the source one.
The problem was partially solved, checking out the
parameters known by the controller, i.e. if the two host are connected to two different switches and the controller switches the source with the
destination address, we can switch it back; the result is that the PacketIn event now has correct information.
If the two host are in the same switch, the problem is solved using a controlled packet flooding, described in section \ref{sec:connectivity}.

\section{Conclusions}\label{sec:conclusion}
% Conclusions are a superbrief summary of what has been done and highlighting of the "take home message"
% ehi
For this course project, the report shows how the demo was made, in particular what problem we encountered
and what we exploited, showing what protocols could be unused in the future. We showed a possible simple data structure
that can allow the controller to be able to know the network, its topology and which route to set, based on the traffic.
Moreover, we showed how to exploit the potentialities of the SDN network, giving particular importance on the
statistics and the path switching capabilities.
The multi-graph solution allowed us to connect the network and, based on its condition, choose which weight to apply
in order to give a Quality of Service based network. With some boards that supports the newer version of Openflow
(i.e. 1.5, the latest), the code can be ported, so it can be used with the Ryu core controller and more statistic can
be exploited. Moreover, the controller can set more strictive rules for the flow control and can get more information from the
switches.
